The abundance of Bullet Groups in $\Lambda$CDM

### 2014-03-10 paper/ms.tex
typos

We estimate the expected distribution of displacements between the two
dominant dark matter peaks in two groups of halos corresponding to
galaxy groups and clusters. We find that 50\% to 60\% of the dark
matter  with circular velocities in the range $300\kms$ to $700\kms$
(groups) show multimodal morphologies with displacements between the
dark matter clumps equal or larger than $133\pm21$\hkpc, which
corresponds to the observational constraint the object SL2S
J08544-0121 located at $z=0.35$ which hass a velocity dispersion of
$644^{+69}_{-102}\kms$. For dark matter halos with circular velocities
larger than $700\kms$ (clusters) this fraction rises to $80\%$ to
$85\%$. Using the same simulation we estimate the dark
matter-to-baryon spatial separation and find that that $0.1\%$ to
$1.0\%$ of the low mass systems should present separations equal or
larger than $87\pm 14$\hkpc corresponding to the same observational
benchmark; for halos with circular velocities larger than $700\kms$
this fraction is in the range of $4\%$ to $10\%$, consistent with
previous studies of dark mattter to baryon separations in
clusters. These results present a weak dependence with redshift. 
\end{abstract}

@@

The Bullet Cluster provided a new kind of observational evidence of
the existece of dark matter. Quantifying the displacement between dark
matter and the dominant baryonic component (hot X-ray emitting gas)
\citep{ForeroRomero2010} has been used to test the CDM paradigm itself
by quantifying the substructure velocity requiered to produce such
displacement \citep{Hayashi2006} to finally estimate the expected
abundance of such events in the Universe.

@@

In this Letter we present prediction for the abundance of group-like
systems that might show a DM-baryon displacement. To this end we use a
N-body cosmological simulation with such a resolution that allows us
to identify multimodal dark matter clumps in the circular velocity
range $300-700$\kms.

This paper is organized as the follows. In Section
\ref{sec:simulation} we present the simulation and the halo catalogs
used in this work. We continue in Section \ref{sec:setup} with the
geometry of the problem at hand and the measurements setup. Next in
Section \ref{sec:results} we present our results to finish with a
discussion and conclussion in Sections \ref{sec:discussion} and
\ref{sec:conclussions}.

@@

We use halo catalogs constructed using the BDM algorithm. [...] \citep{Riebe2013}.

In the snapshots at redshift $z=0.0,0.25, 0.5$ and $1.0$ there are
$XX, XX, XX, XX$ host halos with circular velocities $V_{\rm c}>
300$\kms and $XX, XX, XX, XX$ sub-halos with circular velocities
$V_{\rm c}>75$\kms. These two sets of halos constitute the basis for
our analysis.

For each host halo in the sample we find its most massive
sub-halo. Each pair host/sub-halo is considered as a potential Bullet
Group and is kept for the measurement and analysis described in the
next section.

@@

The bullet-like encounter can be instantaneously described by
quantities the following quantities the circular velocity of the host
and the sub-halo, $V_{\rm c,host}$ and $V_{\rm c,  sub}$; the size of
the host halo $R_{\rm vir}$; the relative position and velocity of the
substructre, $\vec{v}$ and $\vec{r}$; and the angle between the
position and velocity $\mu$.

As a first approximation there are two quantities that are available
from observations of these Bullet-like systems. The projected distance
between two dominant dark matter clumps and the ratio of the
galaxies luminosities associated to them. From the simulation point
of view this can be translated into the 2D projected values of
$||\vec{r}||$, its value relative to the virial radius $D_{\rm off}=
||\vec{r}||_{\rm 2D}/R_{\rm vir}$ and the ratio of the circular velocities
of the two clumps $V_{\rm c,sub}/V_{\rm c,host}$. In a higher degree
of detail, in order to gaint better insight we use
the sub-structure velocity as a fraction of the host's circular
velocity, $||\vec{r}||/V_{\rm c, host}$, as a mesure of the the strenght
of the interaction.  Finally, we also measure the geometry of each
interaction through the values for $\mu$.

All the physical quantities described above can be used to describe
the three main stages in a bullet-like encounter. First, the sub-halo
crosses the virial radius of the host halo starting a head on
collission, $||\vec{r}||/R_{\rm vir}\approx 1$ and
$\mu\approx<0.0$. Second, as the sub-halo crosses for the first time
the center of the host halo $||\vec{r}||/R_{\rm vir} < 1.0$ and
$\mu>0.0$. Third, as the sub-halo reaches apogee and comes back to the
center of the halo $||\vec{r}||/R_{\rm vir} < 1.0$ and $\mu<0.0$. We use
this quantities in Section XXX to fully characterize the different
kind of interactions observed in the Bolshoi simulation.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_1.eps}
\end{center}
\caption{
  Integrated probability distribution for the displacement
  between the host halo and its largest sub-halo. The left panel
  shows the results in terms of the physical displacements while
  the right panel shows the displacements normalized by the virial radius of
  the host halo. The continuous line corrresponds to the halos in the
  group sample $V_{\rm circ,host}>700\kms$ and the dashed lines to the
  sample $V_{\rm circ,host}<700\kms$.
  The vertical lines corresponds to mean estimate separation 
  between the two dark matter clumps in the results reported by
  \citet{Gastaldello} for the group SL2S J08544-0121. Between $50\%$
  to $60\%$ of the groups show a displacement equal or larger than
  this observational benchmark.}
\label{fig:displacement}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_2.eps} 
\end{center}
\caption{2D histogram in the plane $V_{\rm c,sub}/V_{\rm
    c,host}$-$d_{2D}$. The left panel corresponds to groups and the
  right panel to clusters. The circle with error bars corresponds to
  \bullg data reported from \citet{2013A&A...552A..80M} and
  \citet{Gastaldello}. The data used to construct the histograms
  integrates the objects at all redshifts. }
\label{fig:mass_displacement}
\end{figure*}

The main result of this paper is summarized in Figure
\ref{fig:displacement}, it presents the integrated
probability distribution for the displacement between the center of
the host halo and its dominant sub-halo. The left pannel shows
the displacement in physical units and the right pannel as a fraction
of the virial radius of the host halo.

@@

To construct this figure we co-add all the halos in the sample (left,
groups; right, clusters) at all redshifts. We stack the data because
we do not observe any strong time evolution, aditionally this allows
us to increase the signal in each bin. Overplotted there is an circle
with error bars that represents the observational estimates for the
system \bullg using the fraction in velocity dispersion in
the line-of-sigth of the group SL2S SJ08544-0121
($\sigma_{host}=341^{+43}_{-109}$\kms and
$\sigma_{sub}=185^{+30}_{-62}$\kms) reported by
\citet{2013A&A...552A..80M} and the DM displacement inferred from the
data presented by \cite{Gastaldello}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_4.eps}
\end{center}
\caption{x}
\label{fig:geometry}
\end{figure*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{figure_5.eps}
\end{center}
\caption{Integrated probability distribution for the estimated
  baryonic displacements in the group and cluster samples at all redshifts.}
\label{fig:baryonic_displacements}
\end{figure}

These different collision geometries will produce different results in
terms of the displacement between dark matter and baryons. Strictly
speaking, the results we have derived so far, apply to multimodal groups and their
expected separation between the two dominant dark matter clumps but it
should not be interpreted that all have a corresponding DM-baryon
displacement. The systems where the halo is starting to fall into the
host ($\mu\approx-1$, $D_{\rm off}>0.6$) should not present a
detectable DM-baryon displacement.

@@

The results for the integrated distributions for $>d_{2D}^{\rm bar}$
are shown in Figure \ref{fig:baryonic_displacements}. The dashed lines
represented the results for groups and the continous lines correspond
to clusters. As a test of our approach we compare the cluster results
against the analytic fit provided by \cite{ForeroRomero2010}. This fit
reproduces the statistics for the DM-baryon separation found for
clusters more massive than $>10^{14}$\hMsun in a simulation
that included a description for gas with $8$ times the volume of the
Bolshoi Simulation. This fit is valid for separations larger than
$70$\hkpc, beyond that separation we find that it provides a good
description within a factor of $\sim 2$ of

\subsection{Reproducibility of our results}
In order to facilitate the reproducibility and reuse of our results we
have made available all the data and the source code available in a public
repository. [...]

@@

Our estimate is based on large N-body DM-only cosmological
simulation. We estimate the distribution of projected displacements
between the dominant DM clumps in two kinds of systems; groups with circular
velocities $300\kms<V_{\rm c}<700\kms$ and clusters with $V_{\rm
  c}>700$\kms. We report these results at four different redshifts
$z=0.0,0.25,0.5$ and $10$.

We find that a fraction of $XX\pm XX\%$ of the halos in the group
sample can present a displacement equal or larger than the observed displacement
for \bullg. We also present these results in terms of the ratio of the
circular velocities of the sub-halo to the host halo to find that the
system \bullg is found within the expected parameter space spanned by
the simulation.

Next we describe the geometry of the collision by the angle between
the position and velocity vectors of the sub-halo with respect to the
host. This allows us to separate three different instances: the
sub-halo crosses the virial radio of the host, the sub-halo goes
through the center of the host and the secondary
infall after reaching apogee.

From this analysis we estimate a the integrated distribution of the
displacements between the DM and the baryonic component. In this case
we find that a fraction of $XX \pm XX\%$ halos in the group sample
should present displacements equal or larger than those observed for
the Bullet-Group \bullg.

The results we present here can be used as a potential test of
$\Lambda CDM$. (Comments about Gaels results on multimodal groups).

>>>>

We estimate the expected distribution of displacements between the two
dominant dark matter peaks in two groups of halos corresponding to
galaxy groups and clusters. We find that 50\% to 60\% of the dark
matter  with circular velocities in the range $300\kms$ to $700\kms$
(groups) show multimodal morphologies with displacements between the
dark matter clumps equal or larger than $133\pm21$\hkpc, which
corresponds to the observational constraint the object SL2S
J08544-0121 located at $z=0.35$ which has a velocity dispersion of
$644^{+69}_{-102}\kms$. For dark matter halos with circular velocities
larger than $700\kms$ (clusters) this fraction rises to $80\%$ to
$85\%$. Using the same simulation we estimate the dark
matter-to-baryon spatial separation and find that that $0.1\%$ to
$1.0\%$ of the low mass systems should present separations equal or
larger than $87\pm 14$\hkpc corresponding to the same observational
benchmark; for halos with circular velocities larger than $700\kms$
this fraction is in the range of $4\%$ to $10\%$, consistent with
previous studies of dark matter to baryon separations in
clusters. These results present a weak dependence with redshift. 
\end{abstract}

@@

The Bullet Cluster provided a new kind of observational evidence of
the existence of dark matter. Quantifying the displacement between dark
matter and the dominant baryonic component (hot X-ray emitting gas)
\citep{ForeroRomero2010} has been used to test the CDM paradigm itself
by quantifying the substructure velocity required to produce such
displacement \citep{Hayashi2006} to finally estimate the expected
abundance of such events in the Universe.

@@

In this Letter we present prediction for the abundance of group-like
systems that might show a DM-baryon displacement. To this end we use a
N-body cosmological simulation with such a resolution that allows us
to identify multi-modal dark matter clumps in the circular velocity
range $300-700$\kms.

This paper is organized as the follows. In Section
\ref{sec:simulation} we present the simulation and the halo catalogs
used in this work. We continue in Section \ref{sec:setup} with the
geometry of the problem at hand and the measurements setup. Next in
Section \ref{sec:results} we present our results to finish with a
discussion and conclusion in Sections \ref{sec:discussion} and
\ref{sec:conclussions}.

@@

We use halo catalogs constructed using the BDM algorithm. [...] 
All the raw data used in this Letter is available through the
Multidark database \footnote{\texttt{www.multidark.org}}
\citep{Riebe2013}.

In the snapshots at redshift $z=0.0,0.25, 0.5$ and $1.0$ there are
$10041, 10346, 10554,$ and $10382$ host halos with circular velocities
$V_{\rm c}\geq 300$\kms; and $157853, 177331, 195072, 225188$
sub-halos with circular velocities $V_{\rm c}\geq 75$\kms,
respectively. These two sets of halos constitute the basis for our
analysis.

For each host halo we find its most massive sub-halo. Each pair
host/sub-halo is considered as a potential Bullet system and is kept
got the analysis described in the next section.

@@

The bullet-like encounter can be instantaneously described by
quantities the following quantities the circular velocity of the host
and the sub-halo, $V_{\rm c,host}$ and $V_{\rm c,  sub}$; the size of
the host halo $R_{\rm vir}$; the relative position and velocity of the
substructure, $\vec{v}$ and $\vec{r}$; and the angle between the
position and velocity $\mu$.

As a first approximation there are two quantities that are available
from observations of these Bullet-like systems. The projected distance
between two dominant dark matter clumps and the ratio of the
galaxies luminosities associated to them. From the simulation point
of view this can be translated into the 2D projected values of
$||\vec{r}||$, its value relative to the virial radius $D_{\rm off}=
||\vec{r}||_{\rm 2D}/R_{\rm vir}$ and the ratio of the circular velocities
of the two clumps $V_{\rm c,sub}/V_{\rm c,host}$. In a higher degree
of detail, in order to gain better insight we use
the sub-structure velocity as a fraction of the host's circular
velocity, $||\vec{r}||/V_{\rm c, host}$, as a measure of the the strength
of the interaction.  Finally, we also measure the geometry of each
interaction through the values for $\mu$.

All the physical quantities described above can be used to describe
the three main stages in a bullet-like encounter. First, the sub-halo
crosses the virial radius of the host halo starting a head on
collision, $||\vec{r}||/R_{\rm vir}\approx 1$ and
$\mu\approx<0.0$. Second, as the sub-halo crosses for the first time
the center of the host halo $||\vec{r}||/R_{\rm vir} < 1.0$ and
$\mu>0.0$. Third, as the sub-halo reaches apogee and comes back to the
center of the halo $||\vec{r}||/R_{\rm vir} < 1.0$ and $\mu<0.0$. We use
this quantities in Section XXX to fully characterize the different
kind of interactions observed in the Bolshoi simulation.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_1.eps}
\end{center}
\caption{
  Integrated probability distribution for the displacement
  between the host halo and its largest sub-halo. The left panel
  shows the results in terms of the physical displacements while
  the right panel shows the displacements normalized by the virial radius of
  the host halo. The continuous line corresponds to the halos in the
  group sample $V_{\rm circ,host}>700\kms$ and the dashed lines to the
  sample $V_{\rm circ,host}<700\kms$.
  The vertical lines corresponds to mean estimate separation 
  between the two dark matter clumps in the results reported by
  \citet{Gastaldello} for the group SL2S J08544-0121. Between $50\%$
  to $60\%$ of the groups show a displacement equal or larger than
  this observational benchmark.}
\label{fig:displacement}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_2.eps} 
\end{center}
\caption{2D histogram in the plane $V_{\rm c,sub}/V_{\rm
    c,host}$-$d_{2D}$. The left panel corresponds to clusters and the
  right panel to groups. The circle with error bars corresponds to
  \bullg data reported from \citet{2013A&A...552A..80M} and
  \citet{Gastaldello}. The data used to construct the histograms
  integrates the objects at all redshifts. }
\label{fig:mass_displacement}
\end{figure*}

The main result of this paper is summarized in Figure
\ref{fig:displacement}, it presents the integrated
probability distribution for the displacement between the center of
the host halo and its dominant sub-halo. The left panel shows
the displacement in physical units and the right panel as a fraction
of the virial radius of the host halo.

@@

To construct this figure we co-add all the halos in the sample (left,
groups; right, clusters) at all redshifts. We stack the data because
we do not observe any strong time evolution, additionally this allows
us to increase the signal in each bin. Overplotted there is an circle
with error bars that represents the observational estimates for the
system \bullg using the fraction in velocity dispersion in
the line-of-sigth of the group SL2S SJ08544-0121
($\sigma_{host}=341^{+43}_{-109}$\kms and
$\sigma_{sub}=185^{+30}_{-62}$\kms) reported by
\citet{2013A&A...552A..80M} and the DM displacement inferred from the
data presented by \cite{Gastaldello}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_4.eps}
\end{center}
\caption{2D histograms in the plane $\mu$-$D_{\rm off}$. The left
  panel corresponds to clusters and the right panel to groups. The data
  used to construct this histogram includes the halos at all redshifts.}
\label{fig:geometry}
\end{figure*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{figure_5.eps}
\end{center}
\caption{Integrated probability distribution for the estimated
  baryonic displacements in the group and cluster samples. The
  continuous line marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and baryons. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.}
\label{fig:baryonic_displacements}
\end{figure}

These different collision geometries will produce different results in
terms of the displacement between dark matter and baryons. Strictly
speaking, the results we have derived so far, apply to multi-modal groups and their
expected separation between the two dominant dark matter clumps but it
should not be interpreted that all have a corresponding DM-baryon
displacement. The systems where the halo is starting to fall into the
host ($\mu\approx-1$, $D_{\rm off}>0.6$) should not present a
detectable DM-baryon displacement.

@@

The results for the integrated distributions for $>d_{2D}^{\rm bar}$
are shown in Figure \ref{fig:baryonic_displacements}. The dashed lines
represented the results for groups and the continuous lines correspond
to clusters. As a test of our approach we compare the cluster results
against the analytic fit provided by \cite{ForeroRomero2010}. This fit
reproduces the statistics for the DM-baryon separation found for
clusters more massive than $>10^{14}$\hMsun in a simulation
that included a description for gas with $8$ times the volume of the
Bolshoi Simulation. This fit is valid for separations larger than
$70$\hkpc, beyond that separation we find that it provides a good
description within a factor of $\sim 2$ of

\subsection{Reproducibility of our results}
In order to facilitate the reproducibility and reuse of our results we
have made available all the data and the source code available in a public
repository\footnote{\texttt{https://github.com/Fernandez-Trincado/Bullet\_Groups-2014}.}

@@

Our estimate is based on large N-body dark matter only cosmological
simulation. We estimate the distribution of projected displacements
between the dominant DM clumps in two kinds of systems; groups with
circular velocities $300\kms<V_{\rm c}<700\kms$ and clusters with $V_{\rm
  c}>700$\kms. We report these results at four different redshifts
$z=0.0,0.25,0.5$ and $10$.

We find that a fraction of $50\%$-$60\%$ of the halos in the group
sample present displacement equal or larger than the observed displacement
for \bullg. For halos in the cluster sample this fraction increases to
$80\%$-$85\%$.

We also describe the collision geometry by the angle between
the position and velocity vectors of the sub-halo with respect to the
host. This allows us to separate three different instances: the
sub-halo crossing the host virial radius, the sub-halo going
through the host's center and the secondary infall after reaching
apogee.

From this analysis we derive an estimate for the displacement between
the DM and the baryonic component. In the group sample
$0.1\%$-$1.0\%$ of the halos show a displacement equal or larger than 
the  measurements of \bullg by \citep{Gastaldello}. In the cluster
sample this fraction rises to $4\%$-$10\%$.

The results we present here can be used as a potential test of
$\Lambda$CDM. [...]
